---
title: "CSDE 502 Winter 2021, Assignment 9"
author: "Colin Baynes (cbaynes@uw.edu)"
output:
  bookdown::html_document2:
    number_sections: yes
    self_contained: yes
    code_folding: hide
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
  pdf_document:
    number_sections: yes
    toc: yes
    fig_cap: yes
    keep_tex: yes
  html_document:
    toc: yes
    df_print: paged
urlcolor: blue
---

```{r, include = F}
library(captioner)
table_nums <- captioner(prefix = "Table")
figure_nums <- captioner(prefix = "Figure")
# for having DIV tags
knitr::knit_hooks$set(class = function(before, options, envir) {
  if(before){
    sprintf("<div class = '%s'>", options$class)
  }else{
    "</div>"
  }
})

knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

___Explanation___:
This assignment is intended to give you more practice delving into the Add Health data set and in manipulating additional variables. 

___Instructions___: 

1. Make sure your Rmd file has no local file system dependencies (i.e., anyone should be able to recreate the output HTML using only the Rmd source file).
1. Make a copy of this Rmd file and add answers below each question. The code that generated the answers should be included, as well as the complete source code for the document.
1. Change the YAML header above to identify yourself and include contact information.
1. For any tables or figures, include captions and cross-references and any other document automation methods as necessary.
1. Make sure your output HTML file looks appealing to the reader.
1. Upload the final Rmd to your github repository.
1. Download [`assn_id.txt`](http://staff.washington.edu/phurvitz/csde502_winter_2021/assignments/assn_id.txt) and include the URL to your Rmd file on github.com.
1. Create a zip file from your copy of `assn_id.txt` and upload the zip file to the Canvas site for Assignment 9. ___The zip file should contain only the text file. Do not include any additional files in the zip file--everything should be able to run from the file you uploaded to github.com. Please use zip format and not 7z or any other compression/archive format.___


#
__Using the full household roster (you'll need to go back the full raw data source, [21600-0001-Data.dta](http://staff.washington.edu/phurvitz/csde502_winter_2021/data/21600-0001-Data.dta.zip)), create the following variables for each respondent. Document any decisions that you make regarding missing values, definitions, etc. in your narrative as well as in the R code.  Include a frequency tabulation and a histogram of each result.__
```{r, include=F}
#upload full raw data set
setwd('C:/Users/Cbayn/Dropbox/CSDE 502/week 9')
library(readstata13)
df <- read.dta13("21600-0001-Data.dta")
#save in temporary folder and return as a rds object
tmpdir <- Sys.getenv("TEMP")
write.csv(x = df, file = file.path(tmpdir, "df.csv"), row.names = FALSE)
saveRDS(object = df, file = file.path(tmpdir, "df.RDS"))
df_rds <- readRDS(file = file.path(tmpdir, "df.RDS"))
```

__Total number in household__

See `r figure_nums(name = "tcap0", display = "cite")`

_`r figure_nums(name = "tcap0", caption = "Household size of students")`_
```{r, echo=F, warning=F}
df.1 <- subset(df_rds, select = c(H1HR2A, H1HR2B, H1HR2C, H1HR2D, H1HR2E, H1HR2F, H1HR2G, H1HR2H, H1HR2I, H1HR2J, H1HR2K, H1HR2J, H1HR2M, H1HR2N, H1HR2O, H1HR2P, H1HR2Q, H1HR2R, H1HR2S, H1HR2T))
attach(df.1)
library(dplyr)
df.2 <- df.1 %>% mutate_all(funs(type.convert(as.numeric(.))))
library(magrittr)
df.2 %<>% 
    mutate_at(.vars = vars(matches("H1HR2.*")),
              funs(
                  case_when(
                      . == 2 ~ 1,
                      . > 1 ~ 0,
                      TRUE ~ as.numeric(.)
                  )
              )
    )
df.2$total.HH <-rowSums(df.2)
table(df.2$total.HH)
df.2$total.HH<-as.character(df.2$total.HH)
library(ggplot2)
g1 <- ggplot(data = df.2, mapping = aes(x = total.HH)) +
  geom_histogram(stat = "count") +
  scale_x_discrete(limits = c(1:12)) +
  xlab("Total number of household members") 
g1 + ggtitle("Figure 1: Distribution of members of students' households")
```
#

__Number of sisters__

See `r figure_nums(name = "tcap1", display = "cite")`

_`r figure_nums(name = "tcap1", caption = "Number of sisters living with student in same household")`_
```{r, echo=F, warning=F}
df.3 <- subset(df_rds, select = c(H1HR3A, H1HR3B, H1HR3C, H1HR3D, H1HR3E, H1HR3F, H1HR3G, H1HR3H, H1HR3I, H1HR3J, H1HR3K, H1HR3J, H1HR3M, H1HR3N, H1HR3O, H1HR3P, H1HR3Q, H1HR3R, H1HR3S, H1HR3T))
attach(df.3)
df.4 <- df.3 %>% mutate_all(funs(type.convert(as.numeric(.))))

df.4 %<>% 
    mutate_at(.vars = vars(matches("H1HR3.*")),
              funs(
                  case_when(
                      . == 8 ~ 1,
                      . != 8 ~ 0,
                      TRUE ~ as.numeric(.)
                  )
              )
    )
df.4$total.sisters <-rowSums(df.4)
table(df.4$total.sisters)
df.4$total.sisters<-as.character(df.4$total.sisters)
g2 <- ggplot(data = df.4, mapping = aes(x = total.sisters)) +
  geom_histogram(stat = "count") +
  scale_x_discrete(limits = c(0:12)) +
  xlab("Total number of sisters") 
g2 + ggtitle("Figure 2: Distribution of sisters of students that live in the students' households")
```


__Number of brothers__

See `r figure_nums(name = "tcap2", display = "cite")`

_`r figure_nums(name = "tcap2", caption = "Number of brothers living with student in same households")`_
```{r, echo=F, warning=F}
df.3 <- subset(df_rds, select = c(H1HR3A, H1HR3B, H1HR3C, H1HR3D, H1HR3E, H1HR3F, H1HR3G, H1HR3H, H1HR3I, H1HR3J, H1HR3K, H1HR3J, H1HR3M, H1HR3N, H1HR3O, H1HR3P, H1HR3Q, H1HR3R, H1HR3S, H1HR3T))
attach(df.3)
df.5 <- df.3 %>% mutate_all(funs(type.convert(as.numeric(.))))

df.5 %<>% 
    mutate_at(.vars = vars(matches("H1HR3.*")),
              funs(
                  case_when(
                      . == 5 ~ 1,
                      . != 5 ~ 0,
                      TRUE ~ as.numeric(.)
                  )
              )
    )
df.5$total.brothers <-rowSums(df.5)
table(df.5$total.brothers)
df.5$total.brothers<-as.character(df.5$total.brothers)
library(ggplot2)
g3 <- ggplot(data = df.5, mapping = aes(x = total.brothers)) +
  geom_histogram(stat = "count") +
  scale_x_discrete(limits = c(0:12)) +
  xlab("Total number of brothers") 
g3 + ggtitle("Figure 3: Distribution of brothers of students that live in the students' 
             households")
```


__Total number of siblings__

See `r figure_nums(name = "tcap3", display = "cite")`

_`r figure_nums(name = "tcap3", caption = "Number of siblings living with student in same households")`_
```{r, echo=F, warning=F}
df.3 <- subset(df_rds, select = c(H1HR3A, H1HR3B, H1HR3C, H1HR3D, H1HR3E, H1HR3F, H1HR3G, H1HR3H, H1HR3I, H1HR3J, H1HR3K, H1HR3J, H1HR3M, H1HR3N, H1HR3O, H1HR3P, H1HR3Q, H1HR3R, H1HR3S, H1HR3T))
attach(df.3)
df.6 <- df.3 %>% mutate_all(funs(type.convert(as.numeric(.))))

df.6 %<>% 
    mutate_at(.vars = vars(matches("H1HR3.*")),
              funs(
                  case_when(
                      . == 5 ~ 1,
                      . == 8 ~ 1,
                      . < 5 ~ 0,
                      . == 6 ~ 0,
                      . == 7 ~ 0,
                      . > 8 ~ 0,
                                            TRUE ~ as.numeric(.)
                  )
              )
    )
df.6$total.siblings <-rowSums(df.6)
table(df.6$total.siblings)
df.6$total.siblings<-as.character(df.6$total.siblings)
library(ggplot2)
g4 <- ggplot(data = df.6, mapping = aes(x = total.siblings)) +
  geom_histogram(stat = "count") +
  scale_x_discrete(limits = c(0:12)) +
  xlab("Total number of siblings") 
g4+ ggtitle("Figure 4: Distribution of siblings of students that live in the students' households")
```

#

__What proportion of students live with two biological parents? Include the analysis in your R code.__
```{r, warning=F}
df.7 <- subset(df_rds, select = c(H1HR6A, H1HR6B, H1HR6C, H1HR6D, H1HR6E, H1HR6F, H1HR6G, H1HR6H, H1HR6I, H1HR6J, H1HR6K, H1HR6J, H1HR6M, H1HR6N, H1HR6O, H1HR6P, H1HR6Q, H1HR6R, H1HR6S, H1HR6T))
attach(df.3)
df.8 <- df.7 %>% mutate_all(funs(type.convert(as.numeric(.))))

df.8 %<>% 
    mutate_at(.vars = vars(matches("H1HR6.*")),
              funs(
                  case_when(
                      . == 1 ~ 1,
                      . == 7 ~ 1,
                      . == 2 ~ 0,
                      . == 3 ~ 0,
                      . == 4 ~ 0,
                      . == 5 ~ 0,
                      . == 6 ~ 0,
                      . > 7 ~ 0,
                                            TRUE ~ as.numeric(.)
                  )
              )
    )
df.8$total.biopar <-rowSums(df.8)
table(df.8$total.biopar)
3279/6504
```
Approximately 50.4% of students live with two biological parents.


__Calculate the number of household members that are NOT biological mother, biological father, full brother or full sister. Create a contingency table and histogram for this variable.__


```{r, include==F, warning=F}
df.9 <- cbind(df.2$total.HH, df.8$total.biopar)
colnames(df.9) <- c("total.HH", "total.biopar")
df.9<-as.data.frame(df.9)
df.9$total.HH<-as.numeric(df.9$total.HH)
df.9$total.biopar<-as.numeric(df.9$total.biopar)
df.9$total_not_biopar <- df.9$total.HH - df.9$total.biopar
```
See `r figure_nums(name = "tcap4", display = "cite")`

_`r figure_nums(name = "tcap4", caption = "Number of household members living with student that are not biological parents of student")`_
```{r echo=F, warning=F}
#table(df.9$total_not_biopar)
df.9$total.total_not_biopar<-as.character(df.9$total_not_biopar)
library(ggplot2)
g5 <- ggplot(data = df.9, mapping = aes(x = total_not_biopar)) +
  geom_histogram(stat = "count") +
  scale_x_discrete(limits = c(0:15)) +
  xlab("Total number of siblings") 
g5+ ggtitle("Figure 5: Distribution of members of students' households that are not biological parents")
```

See `r table_nums(name = "tcap0", display = "cite")`

_`r table_nums(name = "tcap0", caption = "No. household members of students that are not biological parents")`_

```{r, echo=F, warning=F}
library(kableExtra)
df.9.tab <-
  df.9 %>% 
    group_by(total_not_biopar) %>% 
    summarise(n = n()) %>% 
    mutate(`%` = n / sum(n) * 100) %>% 
    mutate(`%` = `%` %>% round(1)) %>% 
    kable(caption = "Table 1: No. household members of students that are not biological parents") %>% 
    kable_styling(full_width = FALSE, position = "left")
df.9.tab
```
# 

__Source code__
```{r comment=''}
cat(readLines(con = "csde502_2021_assignment09.Rmd"), sep = '\n')
```
